---
title: "UFC Data Exploration"
author: "Isaac Amouzou"
format: html
editor: visual
---

## Libraries

```{r}
library(tidyverse)
```

## Check data structure and missing

```{r}
data <- read_csv("Data/data.csv")

#get rid of paranthesis and replace slashes with . to have proper col names
names(data) <- names(data) %>%
  gsub("\\(|\\)", "", .) %>%   
  gsub("\\\\|/", ".", .) 
```

Variables are stored as numeric or categorical but some could be logical.

```{r}

#numeric columns with 5 or fewer unique values are casted as factor
data <- data %>%
  mutate(across(where(is.numeric), ~ if(length(unique(.)) <= 5) factor(.) else .))

summary_types <- sapply(data, class)

cat("Number of factor columns:", sum(summary_types == "factor"), "\n")
cat("Number of numeric columns:", sum(summary_types %in% c("numeric", "integer")), "\n")
cat("Number of character columns:", sum(summary_types == "character"), "\n")
```

```{r}
perc.missing <- sapply(data,
                          function(x){
                            mean(is.na(x)) * 100})

missing_df <- data.frame(
  perc_missing = perc.missing) %>% 
  arrange(desc(perc_missing))
head(missing_df)  
```

Nothing more than 23.73586% missing so this is good.

```{r}
na.omit(data) %>% dim()

```

3890 rows left when missing is removed (or about 64% of the dataset left)

```{r}
write.csv(na.omit(data), "Data/na_omitted_data.csv", row.names = FALSE)
```

Create dataset with imputed data

```{r}
numeric_cols <- names(data)[sapply(data, is.numeric)]
numeric_missing_cols <- numeric_cols[sapply(data[numeric_cols], function(x) any(is.na(x)))]
```

```{r}
library(mice)

(starttime <- Sys.time())

miters <- 3

#with 5 multiple imputations
imp <- mice(data, m = miters, method = "pmm", seed = 507, printFlag = F)

#get imputed datasets
completed_list <- lapply(1:miters, function(i) complete(imp, i))

#average datasets to get final imputation dataset

average_imputed <- completed_list[[1]]  

for (col in names(average_imputed)) {
  
  if (is.numeric(average_imputed[[col]])) {
    
    sum_col <- 0
    for(i in 1:length(completed_list)){
      sum_col <- sum_col + completed_list[[i]][[col]]
    }
    
    average_imputed[[col]] <- sum_col / length(completed_list)
  }
}

(endttime <- Sys.time())
```

```{r}
endttime - starttime
```

```{r}
data_numeric_imputed <- data
data_numeric_imputed[numeric_missing_cols] <- average_imputed[numeric_missing_cols]
data_numeric_imputed %>% na.omit(.) %>% dim(.)

```

5894/6012 or about 98% of the data retianed.

```{r}
write.csv(na.omit(data_numeric_imputed), "Data/data_numeric_imputed.csv", row.names = FALSE)
```

## Numeric relationships

```{r}
# fix later - tired now

corr_matrix <- data %>%
  select(where(is.numeric)) %>%
  cor(use = "pairwise.complete.obs")

high_corr <- as.data.frame(as.table(corr_matrix)) %>%
  filter(Var1 != Var2, abs(Freq) > 0.6) %>%
  arrange(desc(abs(Freq)))

dim(high_corr)
```

A lot of variables with decent to high correlation variables

# TODO

### 
